---
title: "Seminar 1"
author: Neil Lloyd
date: 20 January 2025
output: html_document
---

## Installation

You may need to `install.packages("rmarkdown")`. 

## Basics of markdown

Check out this link to learn more about the basics of [Markdown](https://www.markdownguide.org/basic-syntax/). 

For example, you can make lists;

- item 1
- item 2

*Note, be sure to add a line-break before and after the list.*

ordered/numbered lists; 

1. number 1
2. number 2

use LaTeX;
$$
  Y = X\beta + u
$$
even, inline $Y = X\beta + u$.


## Adding code

Load `tidyverse` library. Check this [link](https://rmarkdown.rstudio.com/lesson-3.html) to learn how to hide code-blocks and compress messages/warnings.

```{r,include=FALSE}
library(tidyverse)
```

Create graph 
```{r, fig.cap = "My graph"}
car_frame <- mpg
ggplot(data = car_frame) +
  geom_point(mapping=aes(x=displ, y=hwy)) +
  geom_smooth(mapping=aes(x=displ, y=hwy))
```

## Quarto

I would recommend that you investigate the cool things you can do with [Quarto](https://quarto.org/). 

## Git and GitHub

Here's a great [youtube](https://www.youtube.com/watch?v=RGOj5yH7evk) on Git and GitHub.

The basic steps, *I used*:

1. Create a [GitHub](https://github.com/) account. 
2. Install [Git](https://git-scm.com/downloads) on your computer. This will include an app called Git Bash. You can also check out [Git Desktop](https://desktop.github.com/download/). I have found it easier to follow the steps below, which can be executed from Git Bash or the terminal within RStudio. 
3. I typically follow the following steps laid out on this [blog](https://www.freecodecamp.org/news/the-beginners-guide-to-git-github/) under the heading "*Type 2: Work on your project locally then create the repository on GitHub and push it to remote.*"
